---
title: "R-package for calculation of the Nature Index."
author: "Bård Pedersen" 
date: "April 21, 2021"
output: github_document
---


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


# NIcalc

*NIcalc* contains a set of utility functions tailored for working with the Nature Index for Norway. Functions in *NIcalc* may be used to upload new and revised data into the Nature Index database, to harvest and assemble data sets from the database, to use these data sets in calculations of the Nature Index and similar indices, and to summarize and visualize the results.

## Documentation

*NIcalc* contains documentation and explanations for all functions included. There are three vignettes: *NatureIndexCalculation* describes the mathematical framework for calculating the Nature Index and the associated terminology, *objectsInNIcalc* describes the S3 classes defined within *NIcalc*, while *Distributions* describes how one may interact with the Nature Index database in scripts when uploading new and revised data.

## Install by:

```{r, eval = F}
devtools::install_github("NINAnor/NIcalc", build_vignettes = T)
```

## The most important functions

The functions *importDatasetApi()*, *assembleNiObject()*, *calculateIndex()*, *getIndicatorValues()*, *setIndicatorValues()*, and *writeIndicatorValues()* represent the core of *NIcalc*.

#### Updating the Nature Index database with new data

*getIndicatorValues()*, *setIndicatorValues()*, and *writeIndicatorValues()* may be used in scripts for updating the NI database with new or revised measurements. *getIndicatorValues()* retrieves the current values for a given indicator from the NI database as an S3 object of class *indicatorData*, *setIndicatorValues()* updates *indicatorData* objects with new measurements, and *writeIndicatorValues()* posts objects with updated values to the Nature Index database.

#### How to calculate the Nature Index

*importDatasetApi()* harvests data sets from the database. *assembleNiObject()* assembles and structures data into a complete and consistent data set for calculating the Nature Index. *calculateIndex()* calculates the Nature Index and similar indices. It produces an extensive output for each index value to facilitate further analyses of the results.

An example:

```{r, eval = F}
# Not run: 
# Import data set
amphibiaImport <- NIcalc::importDatasetApi(username = "...",password = "...",
                        eco = NULL, indic = c("Småsalamander","Buttsnutefrosk",
                                              "Storsalamander"),
                        year = c("2010","2014","2019"))
# Assemble
amphibiaInput <- NIcalc::assembleNiObject(inputData = amphibiaImport,
                        predefNIunits = c(allArea = T, parts = T, counties = F),
                        indexType = "thematic")
# Calculate index for amphibians 
amphibiaIndex <- NIcalc::calculateIndex(x = amphibiaInput, nsim = 1000,
                        fids = FALSE, tgroups = FALSE, keys = "ignore",
                        w = 0, awbs = TRUE, awBSunit = "Ferskvann")

```

## Contributing to NIcalc

### Questions

If you have questions regarding *NIcalc* please contact bard.pedersen@nina.no. 

### Reporting a bug/error

Bug reports can submitted via email to bard.pedersen@nina.no or you can submit an issue [here](https://github.com/NINAnor/NIcalc/issues)

Please provide a reproducible error so that we can see what went wrong. [Reprex](https://github.com/tidyverse/reprex) is very helpful in this regard.

Please tell us what you expected to achieve and what the error message was. 

### Submitting code 

Pull requests are the best way to propose changes to the code. To do this please follow these steps:

* Fork the repo and create your branch from master.
If you've added code that should be tested, add tests.

* Create a pull request

* Add an issue associated with each pull request!

* Any contributions you make will be under the same License that covers the project. Feel free to contact the maintainers if that's a concern.


